---
title: "Google Data Analytics Capstone"
author: "Antonio Iannopollo"
date: "2025-06-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(lubridate)
library(readr)
library(ggplot2)
```

## Business Task
Analyze FitBit Fitness Tracker Data to identify trends in smart device usage (activity, sleep, heart rate) and provide high-level recommendations for Bellabeat’s marketing strategy for the Leaf tracker, focusing on growth in the global smart device market.

## Data Sources
Source: FitBit Fitness Tracker Data (CC0, Kaggle, via Mobius), from 30 users, including daily activity, minute-level sleep, and second-level heart rate data.

Files Used:

- dailyActivity_merged_m1.csv, m2.csv: Steps, distance, calories.

- minuteSleep_merged_m1.csv, m2.csv: Sleep states (asleep, restless, awake).

- heartrate_seconds_merged_m1.csv, m2.csv: Heart rate.

Note: Original file names were modified by adding '_m1' or '_m2' to distinguish between files with identical names in the dataset.

Limitations: Small sample (30 users), no demographic data, outdated (2016), missing sleep (751 records) and heart rate (919 records) data.

## Data Cleaning and Processing
Activity: Merged dailyActivity_merged_m1.csv and m2.csv, standardized dates (mdy), no duplicates or missing values.

Sleep: Merged minuteSleep_merged_m1.csv and m2.csv, aggregated to daily totals (minutes asleep, restless, awake, time in bed, sleep records), no duplicates.

Heart Rate: Merged heartrate_seconds_merged_m1.csv and m2.csv, aggregated to daily averages, no duplicates.

Merged Data: Combined datasets by Id and Date using left_join.

```{r load-data, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# Load and process data
daily_activity <- bind_rows(read_csv("dailyActivity_merged_m1.csv"), read_csv("dailyActivity_merged_m2.csv")) %>%
  mutate(ActivityDate = mdy(ActivityDate)) %>% rename(Date = ActivityDate)
sleep_combined <- bind_rows(read_csv("minuteSleep_merged_m1.csv"), read_csv("minuteSleep_merged_m2.csv")) %>%
  mutate(date = mdy_hms(date)) %>% mutate(Date = as.Date(date))
sleep_daily <- sleep_combined %>%
  group_by(Id, Date) %>%
  summarize(TotalMinutesAsleep = sum(value == 1, na.rm = TRUE),
            TotalMinutesRestless = sum(value == 2, na.rm = TRUE),
            TotalMinutesAwake = sum(value == 3, na.rm = TRUE),
            TotalTimeInBed = sum(value %in% c(1, 2, 3), na.rm = TRUE),
            SleepRecords = n_distinct(logId)) %>%
  ungroup()
heartrate_combined <- bind_rows(read_csv("heartrate_seconds_merged_m1.csv"), read_csv("heartrate_seconds_merged_m2.csv")) %>%
  mutate(Time = mdy_hms(Time)) %>% mutate(Date = as.Date(Time))
heartrate_daily <- heartrate_combined %>%
  group_by(Id, Date) %>%
  summarize(AvgHeartRate = mean(Value, na.rm = TRUE)) %>%
  ungroup()
combined_data <- daily_activity %>%
  left_join(heartrate_daily, by = c("Id", "Date")) %>%
  left_join(sleep_daily, by = c("Id", "Date"))
```
## Analysis Summary
Activity: Strong positive correlation between steps and calories (r = 0.59, p < 2.2e-16, 1397 records), indicating fitness tracking is a primary use.

Sleep: Weak negative correlation between sleep and steps (r = -0.17, p = 1.591e-05, 646 records), suggesting sleep tracking is used for rest/recovery.

Heart Rate: Weak positive correlation between heart rate and steps (r = 0.23, p = 3.812e-07, 478 records), indicating niche use.

Usage Patterns: All users track activity (100% of records), but only 46% track sleep and 34% track heart rate, reflecting inconsistent usage.

```{r summary-stats, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
## Summary statistics
activity_summary <- daily_activity %>%
  summarize(AvgSteps = mean(TotalSteps, na.rm = TRUE), AvgCalories = mean(Calories, na.rm = TRUE))
sleep_summary <- sleep_daily %>%
  summarize(AvgSleepMinutes = mean(TotalMinutesAsleep, na.rm = TRUE), AvgTimeInBed = mean(TotalTimeInBed, na.rm = TRUE))
heartrate_summary <- heartrate_daily %>%
  summarize(AvgHeartRate = mean(AvgHeartRate, na.rm = TRUE))
usage_summary <- combined_data %>%
  summarize(Users_Activity = n_distinct(Id[!is.na(TotalSteps)]), Users_Sleep = n_distinct(Id[!is.na(TotalMinutesAsleep)]),
            Users_HeartRate = n_distinct(Id[!is.na(AvgHeartRate)]))
print(activity_summary)
print(sleep_summary)
print(heartrate_summary)
print(usage_summary)
```

## Visualizations

```{r plot-steps-calories, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(combined_data, aes(x = TotalSteps, y = Calories)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Strong Correlation: Steps vs. Calories Burned", x = "Total Steps", y = "Calories",
       caption = "r = 0.59, p < 2.2e-16 (1397 records)") +
  theme_minimal()
```

```{r plot-sleep-steps, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(combined_data, aes(x = TotalMinutesAsleep, y = TotalSteps)) +
  geom_point(color = "green", alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Weak Negative Correlation: Sleep vs. Steps", x = "Minutes Asleep", y = "Total Steps",
       caption = "r = -0.17, p = 1.591e-05 (646 records)") +
  theme_minimal()
```

```{r plot-heartrate-steps, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(combined_data, aes(x = AvgHeartRate, y = TotalSteps)) +
  geom_point(color = "purple", alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Weak Positive Correlation: Heart Rate vs. Steps", x = "Average Heart Rate (bpm)", y = "Total Steps",
       caption = "r = 0.23, p = 3.812e-07 (478 records)") +
  theme_minimal()
```

```{r feature, echo=FALSE, message=FALSE, warning=FALSE}
percent_data <- data.frame(
  Feature = c("Activity", "Sleep", "Heart Rate"),
  Percent = c(100, mean(!is.na(combined_data$TotalMinutesAsleep)) * 100, mean(!is.na(combined_data$AvgHeartRate)) * 100)
)
```
```{r feature-usage, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(percent_data, aes(x = Feature, y = Percent, fill = Feature)) +
  geom_bar(stat = "identity") +
  labs(title = "Percentage of Records with Data by Feature", x = "Feature", y = "Percentage of Records (%)",
       caption = "Activity: 1397 records, Sleep: 646 records, Heart Rate: 478 records") +
  theme_minimal() +
  scale_fill_manual(values = c("Activity" = "blue", "Sleep" = "green", "Heart Rate" = "purple"))
```

## Recommendations for Bellabeat
Emphasize Activity Tracking: Market Leaf as a stylish fitness tracker: “Track steps and calories with Leaf’s elegant design.” Use Instagram/YouTube ads.

Promote Sleep Tracking with Education: Highlight sleep insights and encourage nighttime wear: “Discover better rest with Leaf.”

Target Heart Rate as Premium Feature: Position heart rate for fitness/stress monitoring: “Monitor your heart rate with Leaf.”

Address Usage Barriers: Promote Leaf’s versatility (bracelet, necklace, clip) and app prompts to increase sleep/heart rate tracking.

Leverage Digital Marketing: Run video ads on Google, Instagram, YouTube for key dates (e.g., New Year’s).

Collect Proprietary Data: Gather female-specific data to validate trends for Bellabeat’s audience.